---
title: "Problem 5"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
library(GGally)
library(ISLR)
library(ggplot2)
library(ggfortify)
library(MASS)
library(class)
library(pROC)
library(plotROC)
library(caret)
library(boot)
```


```{r importing_data, echo=FALSE}
id <- "19auu8YlUJJJUsZY8JZfsCTWzDm6doE7C" # google file ID
d.bodyfat <- read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download",
id), header = T)
```


### Problem 5a)
```{r 5a}
#Fit the regression model with bodyfat as response, and age, weight and bmi as predictors
bodyfat.lm <- lm(bodyfat ~ age+weight+bmi, data=d.bodyfat)

#Display the parameters of the model
summary(bodyfat.lm)
```
In the code cell above, a linear regression model has been used to model bodyfat as the response with age, weight and bmi as predictor variables. R-squared can be read from the table above, and one gets $R^2=$ `r signif(summary(bodyfat.lm)$r.squared,3)`.


### Problem 5b)

```{r 5b}
set.seed(4268)#For consistent results

#Function fits a linear regression model and returns R^2
boot.fnc <- function(formula ,data , indices){
  d <- data[indices, ]
  fitBoot <- lm(formula, data=d)
  return(summary(fitBoot)$r.square)
}

#Generating 1000 bootstrap samples
rsquared.boot <- boot(data=d.bodyfat, statistic=boot.fnc,R=1000,
              formula=bodyfat~age+weight+bmi)

#Display the result of the bootstraping
rsquared.boot

#Plotting the distribution of R^2
plot(rsquared.boot)

#Calculates confidence interval for the R^2 estimate
boot.ci(rsquared.boot, type='bca')
```
```{r echo=FALSE}
#Used for extracing the confidence interval
el = boot.ci(rsquared.boot, type='bca')
```


**iii)** As one can see in from the "Bootstrap Statistics" above, the standard error for $R^2$ is `r signif(sqrt(var(rsquared.boot$t)),3)`. A $95\%$ confidence interval for $R^2$ is \
`r signif(el$bca[4],3)` $<R^2<$ `r signif(el$bca[5],3)`.


**iv)** This indicates that what seems like a certain value, actually is a point-estimate and is surprisingly uncertain. One must therefore have this uncertainty in mind  when working with the $R^2$-value.




