---
title: "Problem 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
library(GGally)
library(ISLR)
library(ggplot2)
library(ggfortify)
library(MASS)
library(class)
library(pROC)
library(plotROC)
library(caret)
```



```{r importing data, echo=FALSE}
id <- "1yYlEl5gYY3BEtJ4d7KWaFGIOEweJIn__" # google file ID
d.corona <- read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download",
                             id), header = T)
```

## Problem 2a)


Here one can see a table with the number of deceased (1) and non-deceased (0) in the whole dataset. 
```{r 2a.i, echo=TRUE}
table(d.corona[,1])
```
Here one can see a table with the number of males and females in each country.
```{r 2a.ii, echo=TRUE}
table(d.corona[,c(2,4)])

```

Here one can see a table with the number of deceased and non-deceased for each sex.
```{r 2a.iii, echo=TRUE}
table(d.corona[c(1,2)])
```

Here one can see a table with the number of deceased and non-deceased for each sex in France.
```{r 2a.iv, echo=TRUE}
table(subset(d.corona,country=="France")[,c(1,2)])
```



## Problem 2b)

When choosing a model for this problem, several considerations were made. First, a linear model gave negative probabilities. Second, the questions in this problem would be hard to answer using LDA, QDA or KNN, due to inference being important. Therefore we chose a logistic regression model. The summary and an analysis of the variance with the chi-squared-test can be seen under.

```{r 2b_building_model}
deceased.logm  <- glm(deceased ~ . ,data=d.corona,family="binomial") #Training a log.regr. model on dataset.
summary(deceased.logm)
anova(deceased.logm, test="Chisq")

```


**i)**

```{r 2b.i}
koreanMan75 <- data.frame(sex="male", age=75, country="Korea") #Creating a 74 yrs old Korean man.
koreanMan75.prediction <- predict.glm(deceased.logm, koreanMan75, type="response") #Predicting
koreanMan75.prediction

```
The logistic regression model predicts a `r signif(koreanMan75.prediction,3)` chance of dying for a $75$ year old korean man.



**ii)**
The estimated predictor for the sexmale-variable is `r signif(summary(deceased.logm)$coefficients[2,1],3)` with p-value `r signif(summary(deceased.logm)$coefficients[2,4],3)`. Since the p-value is quite small, there is most likely a relation between dying and one's sex, and since the sexmale-predictor is positive, this indicates a higher probability to die for a male than a female.

**iii)**
The p-value of the Chi-squared test for the country-variable is `r signif(anova(deceased.logm, test="Chisq")$"Pr(>Chi)"[4],3)`. This is a pretty small p-value, which indicates a relation between dying and one's country of residence.


**iv)**
Let $A$ be a person and $A_{+10}$ be an identical person but whom is ten years older. Then for a logistic regression model, the following holds:
$$
\frac{Odds(A_{+10})}{Odds(A)} = e^{10\beta_{age}}.
$$
Hence, the odds to die increase by a factor of ${\left(e^{\beta_{age}}\right)}^{10} \approx 1.31$ for an identical person but whom is $10$ years older.



## Problem 2c)

To find out whether two covariates influence each other, one can create a logistic regression model with interaction terms between the two covariates of interest. In order not to waste information, all covariates are included in the models.

**i)**
```{r 2c.i}
deceased_ageSex.logm <- lm(deceased ~ country+age*sex, data=d.corona)
summary(deceased_ageSex.logm) #summary(deceased_ageSex.logm)$coefficients[7,4]
```

The p-value of the interaction term between age and sex is `r signif(summary(deceased_ageSex.logm)$coefficients[7,4],3)`, which gives a large probability for the interaction term to not have any effect on the probability to decease. It is not unlikely that age is a greater risk factor for men than for women, but the the p-value is too large to make a conclusion.



**ii)**

```{r 2c.ii}
deceased_ageEthnic.logm <- lm(deceased ~ sex+age*country, data=d.corona)
summary(deceased_ageEthnic.logm)

```

The p-value for the interaction term between age and countryindonesia is `r signif(summary(deceased_ageEthnic.logm)$coefficients[7,4],3)`, which is very low. Since countryFrance is the reference category and the estimate for $\beta_{age:countryindonesia}$ is negative, this p-value indicates that age is a greater risk for the French population than for the Indonesian population.


## Problem 2d)
**i) True, ii) True, iii) True, iv) False**

